---
title: "January 30"
output: html_notebook
---

Intro: slide 167-169

# Packages

Install any of these that you don't have already:

```{r}
options(tidyverse.quiet=T) # no messages!
library(tidyverse)
library(survival) # for analysis
library(survminer) # for graphs
```

# Dancing data

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/dancing.txt"
dance=read_table(my_url)
dance
```

- Does time spent dancing before quitting depend on treatment (visit to dance competition) or age?
- Some of the women quit dancing at the #months shown (`quit=1`), some were still dancing the last we 
knew (`quit=0`).

- Event of interest "quit dancing"
- Women who were never observed to quit called "censored".

- Model has to accommodate censored individuals: there is information there, particularly if observed a 
long time and event still hasn't happened.

## Modelling and response variable

Response variable first:

```{r}
mth=with(dance,Surv(Months,Quit==1))
mth
```

```{r}
dance %>% mutate(y=mth)
```


```{r}
cbind(mth, dance)
```

that thing `mth`:

```{r}
print.default(mth)
```

*actually* has two columns, so that this doesn't work:

```{r}
dance %>% mutate(mth=mth)
```

## Cox proportional-hazards model 

predict "survival time" (time until quitting dancing) from treatment and age, allowing for fact
that some women never observed to quit. 

We use $\alpha=0.10$, since data set is small.

```{r}
dance.1=coxph(mth~Treatment+Age,data=dance)
summary(dance.1)
```

Tests at bottom: *something* predicts time until quitting dancing.

Both treatment and age have a significant effect at $\alpha=0.10$. What kind of effect? Later.

## Model checking

"Martingale residuals": look for no pattern with fitted values.

```{r}
ggcoxdiagnostics(dance.1)
```

Average residual is zero all way across: no problem.

Martingale residuals can go very negative, so don't look at normal quantile plot of them.

## Predictions

To understand effect of age and treatment, do predictions and plot them.

Combos of age 20 and 40, dance competition and not (1 and 0):

```{r}
ages=c(20,40)
treatments=c(0,1)
new=crossing(Treatment=treatments, Age=ages)
View(new)
```

In a regression, run `predict`. In survival analysis, run `survfit`:

```{r}
s=survfit(dance.1,new,data=dance) # model, new data, original data
summary(s)
```

the four columns `survival` are the four combinations in the order they were in `new`.

Easier: make a graph:

```{r}
ggsurvplot(s,conf.int=F)
```

It is better to have a higher chance of continuing to dance for longer (at least if you own a dance studio!) So being up and to the right is better (purple and not red).

Which of the strata is which? Same order as in `new`:

```{r}
new
```

Purple stratum is #4, treatment 1 (going to dance comp) and age 40. These women continue to dance for longest.

Age effect is that older women continue to dance for longer, keeping treatment fixed (green better than red, purple better than  blue).

Treatment effect is that women who go to the dance competition continue to dance for longer, keeping age fixed (blue better than red, purple better than green).

Go back to `summary` output:

```{r}
summary(dance.1)
```

The coefficients are on the scale of "hazard of event": if positive, makes the event more likely to happen sooner, if negative, event is less likely to happen sooner (more likely to take a long time).

Here, coefficients are both negative. That means:

- a woman who goes to the dance competition is less likely to quit soon (more likely to dance for a long time)
- a woman who is older is less likely to quit soon (more likely to dance longer).


# Lung cancer example

The variables: slide 183, all quantitative (except for `inst`, a label, and `status`, `sex`, categorical variables with 2 levels treated as numbers).

```{r}
lung
```

## Remove missing values

There are missing values. How many?

```{r}
summary(lung)
```

Some variables have a lot.

Get rid of cases with missing values

```{r}
(lung %>% drop_na() -> lung.complete)
```

Any missing values left?

```{r}
summary(lung.complete)
```

No. We are good.

## Create response variable and fit Cox model

Create response variable. Use `time` (survival time) and `status`. Check data description to see that 2 means "event happened" (death). These times are days.

```{r}
y=with(lung.complete, Surv(time, status==2))
y
```

As a check, usually the longer survival times go with survival/censoring:

```{r}
ggplot(lung.complete, aes(x=factor(status), y=time))+geom_boxplot()
```

1 was lived, 2 was died.

## Cox model

`inst` is label for "institution", ignore. 

To start, use all the other variables we haven't used as explanatory.
This means using everything except `inst`, `time`, `status`.

```{r}
lung.1=coxph(y~.-inst-time-status, data=lung.complete)
summary(lung.1)
```

There are some significant variables and some non-significant ones.

Suggests `step` with backward elimination to get a better model:

```{r}
lung.2=step(lung.1, test="Chisq",trace=1)
```

```{r}
summary(lung.2)
```

I make the call that I'm only going to keep `sex` and `ph.ecog` and take out the other three:

```{r}
lung.3=update(lung.2, .~.-ph.karno-pat.karno-wt.loss)
summary(lung.3)
```

Interpretation:
- a larger `sex` number (female=2) is less likely to die sooner
- a larger score on `ph.ecog` scale means *more* likely to die sooner.

## Should I have taken out all those variables?

```{r}
anova(lung.3, lung.1)
```

yes, I was justified in removing them all.

## Predictions and graph

Make combinations of sexes and `ph.ecog` scores:

```{r}
sexes=c(1,2)
ph.ecogs=0:3
lung.new=crossing(sex=sexes,ph.ecog=ph.ecogs)
View(lung.new)
```

Predict:

```{r}
s=survfit(lung.3, lung.new, data=lung.complete)
```

Plot them:

```{r}
ggsurvplot(s,conf.int=F)
```

Best survival is light blue curve, stratum 5: females with `ph.ecog` score 0.

Worst survival is green curve, stratum 4: males with `ph.ecog` score 3.

Effect of `ph.ecog` score: compare eg. for females, strata 5 through 8: light blue, dark blue, purple, pink: increasing score associated with worse survival.

Effect of `sex`: compare eg. strata 1 (males, red) and 5 (females, light blue): females have better survival.

```{r}
summary(lung.3)
```


## Martingale residuals

```{r}
ggcoxdiagnostics(lung.3)
```

Grey envelope includes 0 all the way across, so I call this good.

# When the Cox model fails

Invent some data:

```{r}
age=seq(20,60,5)
survtime=c(10,12,11,21,15,20,8,9,11)
stat=c(1,1,1,1,0,1,1,1,1)
d=tibble(age,survtime,stat)
y=with(d,Surv(survtime,stat==1))
cbind(y,d)
```

Survival time 15 was actually censored, so would have been larger if survival time observed.

Fit model:

```{r}
y.1=coxph(y~age,data=d)
summary(y.1)
```

Model and effect of age *not significant*.

Look at martingale residual plot:

```{r}
ggcoxdiagnostics(y.1)
```

Down and up quadratic trend. Suggests adding squared term in age:

```{r}
y.2=coxph(y~age+I(age^2),data=d)
summary(y.2)
```

Squared term almost significant, and

```{r}
ggcoxdiagnostics(y.2)
```

martingale residual plot improved.

What do predictions look like?

```{r}
new=tibble(age=seq(20,60,10))
new
```

```{r}
s=survfit(y.2, new, data=d)
```

```{r}
ggsurvplot(s,conf.int=F)
```

```{r}
new
```

Best survival associated with stratum 3, middling age 40.

Much worse survival for two extreme ages 20 and 60.