---
title: "February 27"
output: html_notebook
---

## Packages

```{r}
library(tidyverse)
library(broom)
library(car)
library(lme4) # for mixed models repeated measures, later
```

## Analysis of covariance

slide 270 for background, 271 for data

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/ancova.txt"
prepost=read_delim(my_url," ")
prepost
```

```{r}
ggplot(prepost,aes(x=before,y=after,colour=drug))+
  geom_point()
```

compare before and after means of each drug

```{r}
prepost %>% group_by(drug) %>%
summarize(before_mean=mean(before),
          after_mean=mean(after))
```

analysis with interaction

```{r}
prepost.1=lm(after~before*drug,data=prepost)
anova(prepost.1)
```

interaction not significant, remove (later), but first, predictions:

```{r}
new=crossing(
before=c(5,15,25),
drug=c("a","b"))
new
```

and then predict:

```{r}
pred=predict(prepost.1,new)
preds=bind_cols(new,pred=pred)
preds
```

plot data with predictions

```{r}
ggplot(prepost, aes(x=before,y=after,colour=drug))+
  geom_point()+
  geom_line(data=preds,aes(y=pred))
```

lines are approx parallel (given amount of variability), supporting lack of interaction.

Or, more simply:

```{r}
ggplot(prepost,
aes(x=before,y=after,colour=drug))+
  geom_point()+
  geom_smooth(method="lm", se=F)
```

In a moment, though, need more complicated way.

Take out interaction:

```{r}
prepost.2=update(prepost.1,.~.-before:drug)
anova(prepost.2)
```

Strong main effects of `before` and `drug`.

Do predictions from this model:

```{r}
pred=predict(prepost.2,new)
preds=bind_cols(new,pred=pred)
preds
```

Each increase of 10 in before score results in 8.3 in predicted after score,
*the same for both drugs*. (Implication of no interaction.)

Plot again, code as before:

```{r}
ggplot(prepost,
  aes(x=before,y=after,colour=drug))+
  geom_point()+
  geom_line(data=preds,aes(y=pred))
```

This time, lines are *exactly* parallel (model says they must be).

Different look at model output, as "regression":

```{r}
summary(prepost.2)
```


```{r}
summary(prepost.1)
```

slide 286 for discussion

## Multivariate analysis of variance (MANOVA)

slide 289

seed yields and weights:

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/manova1.txt"
hilo=read_delim(my_url," ")
hilo
```

boxplots: yield by fertilizer:

```{r}
ggplot(hilo,aes(x=fertilizer,y=yield))+geom_boxplot()
```

and weight by fertilizer

```{r}
ggplot(hilo,aes(x=fertilizer,y=weight))+geom_boxplot()
```

Not convincing evidence of difference on either plot.

One-way ANOVAs:

```{r}
hilo.y=aov(yield~fertilizer,data=hilo)
summary(hilo.y)
```

and

```{r}
hilo.w=aov(weight~fertilizer,data=hilo)
summary(hilo.w)
```

Neither significant. 

But, a better plot, of the two responses against each other labelled by fertilizer:

```{r}
g=ggplot(hilo,aes(x=yield,y=weight,
  colour=fertilizer))+geom_point()
g
```

I add a line to this, for reasons that will become clear:

```{r}
d=tribble(
  ~line_x, ~line_y,
  31, 14,
  38, 10)
d
g+geom_line(data=d,aes(x=line_x,y=line_y,colour=NULL))
```

slide 298

Is there an effect of fertilizer on yield and weight combined? This is what MANOVA finds out.

```{r}
response=with(hilo,cbind(yield,weight))

hilo.1=manova(response~fertilizer, data=hilo)
summary(hilo.1)
```



Neither yield or weight individually are affected by fertilizer, but *in combination* they are.

slide 300

Another way to do the same thing, from package `car`:

```{r}
hilo.2=lm(response~fertilizer,data=hilo)
Manova(hilo.2)
```

with the same result. (We have to do repeated measures, later, this way, but you can do MANOVA either way.)



```{r}
summary(Manova(hilo.2))
```
  

Peanuts example slide 303

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/peanuts.txt"
(peanuts.orig=read_delim(my_url," "))
```

Location and variety are really categorical; `y`, `smk`, `w` are quantitative responses.

```{r}
peanuts.orig %>%
  mutate(location=factor(location),
         variety=factor(variety)) -> peanuts
peanuts
```

create response

```{r}
response=with(peanuts,cbind(y,smk,w)) 
class(response)
response
```

Analysis uses `Manova`:

```{r}
peanuts.1=lm(response~location*variety,data=peanuts)
Manova(peanuts.1)
```

or not, as you prefer:

```{r}
peanuts.3=manova(response~location*variety, data=peanuts)
summary(peanuts.3)
```

Interaction term is almost significant, so there might be a dependence of the 3-variable response on location-variety combination.

Understanding that requires a 3D plot or discriminant analysis (later). 

More detail:

```{r}
summary(Manova(peanuts.1))
```


## Repeated measures

slide 307

Histamine in dogs:

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/dogs.txt"
dogs=read_table(my_url)
dogs
```

Four response columns (histamine levels at different times).

Going the `Manova` way for analysis:

```{r}
response=with(dogs,cbind(lh0,lh1,lh3,lh5))
dogs.1=lm(response~drug,data=dogs)
```

We have some extra setup to express that these are the same thing measured at different times:

```{r}
response
times=colnames(response)
times.df=data.frame(times)
```

What are these?

```{r}
times
```

and

```{r}
times.df
```

note that `times` has gotten turned into a factor. `tibble` doesn't work (instead of `data.frame`) because it doesn't create factors (unless you make it do so).

Then the analysis

```{r}
Manova(dogs.1, idata=times.df, idesign=~times)
```

The effect of drug is different at different times.

To investigate what, we can draw a graph. But we have wide format:

```{r}
dogs
```

and we want all the histamine levels in *one* column:

```{r}
dogs %>% gather(timex,lh,lh0:lh5)
```

Can we get the times as *numbers*?

```{r}
dogs %>% gather(timex,lh,lh0:lh5) %>% 
  mutate(time=parse_number(timex)) -> dogs.long
dogs.long
```

Make an interaction plot of log-histamine against time, labelled by drug:

```{r}
ggplot(dogs.long,aes(x=time, y=lh ,colour=drug, group=drug))+
  stat_summary(fun.y=mean,geom="point")+
  stat_summary(fun.y=mean,geom="line")
```

This shows that log-histamine levels are similar at time zero, but for all times after that, trimethaphan log-histamine levels are higher.

I could add the data:

```{r}
ggplot(dogs.long,aes(x=time, y=lh, colour=drug, group=drug))+
  stat_summary(fun.y=mean,geom="line")+
  geom_point()
```

The log-histamine levels really are all mixed up at time zero, and the blue points really are higher than the red ones after that. This explains the significant interaction. 

Time zero really ought not to be included in the analysis, because there is no time for the drugs to have an effect. (There is usually a value noted at time zero so that we can confirm the drugs were not different then.) 

So let's go back to the original data frame and exclude time zero. This means redefining the response and following through:

```{r}
response=with(dogs,cbind(lh1, lh3, lh5)) # excluding time zero
dogs.2=lm(response~drug, data=dogs)
times=colnames(response)
times
times.df=data.frame(times)
times.df
Manova(dogs.2,idata=times.df, idesign=~times)
```

The interaction is no longer significant.

Cannot take it out (interaction of "within-subject" factor `times` and "between-subject" factor `drug`). So ignore it, because it is non-significant, and interpret the main effects: an effect of time, but an only marginally significant effect of drug. 

A better plot for repeated-measures data is to join measurements on the same individual: a "spaghetti plot", thus:

```{r}
ggplot(dogs.long,aes(x=time, y=lh, colour=drug, group=dog)) +
  geom_point()+geom_line()
```

