---
title: "March 27"
output: html_notebook
---

## packages

```{r}
library(tidyverse)
library(ggbiplot)
```


## principal components from correlation matrix

store in file and read in:

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/cov.txt"
mat=read_table(my_url,col_names=F)
mat
```

correlations between three variables X1-X3.

```{r}
mat %>% as.matrix() %>%
princomp(covmat=.) -> mat.pc
summary(mat.pc)
```

there is one component, which we can also see with a screeplot:

```{r}
ggscreeplot(mat.pc)
```

Clear elbow at 2, so one component.

Compare original correlation matrix:

```{r}
mat
```

with loadings

```{r}
mat.pc$loadings
```

component 1 is large when X1 and X2 are large, X3 small. Small for opposite.

Correlation matrix says that one of these is likely to happen, so data are basically one-dimensional.

see slide 579.

## factor analysis

slide 581

### children test scores

slide 583

read in data (correlation matrix)

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/rex2.txt"
kids = read_delim(my_url," ")
kids
```

first run principal components to get scree plot

```{r}
kids %>%
  select_if(is.numeric) %>%
  as.matrix() %>%
  princomp(covmat=.) -> kids.pc
ggscreeplot(kids.pc)
```

Elbow at 3, two components / factors.

```{r}
kids.pc$loadings
```

- first component has some of everything
- second component mainly `add` and `dots`
- not clear

  slide 587
  
```{r}
kids %>%
  select_if(is.numeric) %>%
  as.matrix() -> km
km2=list(cov=km,n.obs=145)
kids.f2=factanal(factors=2,covmat=km2)
```

Uniquenesses (slide 589)

```{r}
kids.f2$uniquenesses
```

Loadings

```{r}
kids.f2$loadings
```

slide 590

Are two factors enough?

```{r}
kids.f2$PVAL
```

2 factors not rejected, so enough.

Would  1 factor have been enough?

```{r}
kids.f1=factanal(factors=1,covmat=km2)
kids.f1$PVAL
```

No: we need 2 factors.

### track running records revisited

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/men_track_field.txt"
track=read_table(my_url)

my_url="http://www.utsc.utoronto.ca/~butler/d29/isocodes.csv"
iso=read_csv(my_url)

track_num = track %>% select_if(is.numeric)
track.pc=princomp(track_num,cor=T)
ggbiplot(track.pc,labels=track$country)
```


principal component dimensions are good-bad (left-right), sprinting-distance (up-down).

What happens when we do factor analysis? Slide 594.

```{r}
track %>% select_if(is.numeric) %>%
factanal(2,scores="r") -> track.f
```

Biplot (not ggplot):

```{r fig.height=12, fig.width=12}
biplot(track.f$scores,track.f$loadings, xlabs=track$country)
```

100m points up, and marathon points across. Suggests two dimensions will be more clearly sprinting and distance running. Note "rotation" of variables on biplot.

```{r}
track.f$loadings
```

factor 1, highesy loadings on long distance, factor 2 highest loadings on sprinting.

Which countries are best at distance running or sprinting? Most negative scores on factor 1, 2 respectively.

Make data frame with  country abbreviations and factor scores, and add country names. (Compare what happens if I use `cbind`.)

```{r}
data.frame(country=track$country, track.f$scores) %>% 
  left_join(iso,by=c("country"="ISO2")) -> d
d
```

best distance-running countries:

```{r}
d %>% arrange(Factor1)
```

best sprinting countries:

```{r}
d %>% arrange(Factor2)
```

### BEM sex role inventory

slide 602-603.

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/factor.txt"
bem=read_tsv(my_url)
bem
```

Traits have abbreviated names. Scores on 7-point scale: 1=strongly disagree, 7=strongly agree.

Side track: make a lookup table. First need all (distinct) short names:

```{r}
bem %>% gather(trait, score, -subno) %>% distinct(trait) %>% write_csv("traits.csv")
```

then open `traits.csv` with spreadsheet, add new column with full traits in `trait_full`.

then read this back in:

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/traits.csv"
traits=read_csv(my_url)
traits
```






use principal components to decide how many factors:

```{r}
bem %>% select(-subno) %>% princomp(cor=T) -> bem.pc
ggscreeplot(bem.pc)
```

Hard to see. Looking for a reasonably small number of factors:

```{r}
ggscreeplot(bem.pc)+scale_x_continuous(limits=c(0,8))
```

Maybe elbows at 3 (2 factors) or 6 (5 factors).

But:

```{r}
summary(bem.pc)
```

2 factors explains only 29% of variability, and 5 only 43%. We are not going to be able to do so well here.

Biplot:

```{r fig.height=12, fig.width=12}
ggbiplot(bem.pc,alpha=0.3)
```

A lot of variables point to 1 o'clock or 4 o'clock. A rotation might point them up-down or left-right.

2 factors (may be bad):

```{r}
bem.2 = bem %>% select(-subno) %>% factanal(factors=2, scores="r")
bem.2
```

Did we rotate?

```{r fig.height=12, fig.width=12}
biplot(bem.2$scores,bem.2$loadings, xlabs=bem$subno)
```

Most of the red arrows now go up or right, so factors ought to have (somewhat) clearer structure.


What is factor 1 made of?

Trickiness first:

```{r}
bem.2$loadings
class(bem.2$loadings) # not a data frame
bem.2$loadings %>% unclass() %>% as.data.frame() %>% rownames_to_column("trait") -> bem.loadings
bem.loadings
```

Look up full trait names:

```{r}
(bem.loadings %>% left_join(traits) -> bem.loadings)
```


```{r}
bem.loadings %>% arrange(desc(abs(Factor1)))
```

and factor 2?

```{r}
bem.loadings %>% arrange(desc(abs(Factor2)))
```

Clearly two very different types of traits.

Go back to biplot, now looking for unusual *individuals*. The numbers are subject numbers in `subno`:

```{r fig.height=12, fig.width=12}
biplot(bem.2$scores,bem.2$loadings, xlabs=bem$subno)
```

On factor 2, 755 high and 534 low:

```{r}
bem %>% filter(subno %in% c(755, 534)) %>% gather(trait, score, -subno) %>% spread(subno, score) -> extreme_2
extreme_2
```

but we need to look at the columns *which load highly on factor 2*:

```{r}
bem.loadings
```

```{r}
extreme_2 %>% left_join(bem.loadings) %>% arrange(desc(abs(Factor2)))
```

Subject 755 scores highly on these traits (and even low on "masculine" with negative loading!)
Subject 534 scores at least relatively low on these traits.

Repeat with factor 1: 404 high and 708 low:

```{r}
bem %>% filter(subno %in% c(404, 708)) %>% gather(trait, score, -subno) %>% spread(subno, score) %>% 
  left_join(bem.loadings) %>% arrange(desc(abs(Factor1)))
```

subject 404 is a high scorer on everything (and a low scorer on "shy" with negative loading)
subject 708 is mainly a low scorer on these.

Try a crazy number of factors like 15(!)

```{r}
bem %>% select(-subno) %>%
  factanal(factors=15) -> bem.15
```

and then see what each factor has in it. Loadings first, with full trait names

```{r}
bem.15$loadings %>% unclass() %>% as.data.frame() %>% rownames_to_column("trait") %>% 
  left_join(traits) -> bem.loadings
bem.loadings
```

```{r}
bem.loadings %>% gather(factor, score, starts_with("Factor")) -> loadings.long
loadings.long
```

```{r}
high_loadings=function(factor_name) {
  loadings.long %>% filter(factor==factor_name) %>% 
  arrange(desc(abs(score)))
}
```

then do it for any factor

```{r}
factor_name="Factor15"
high_loadings(factor_name)
```

but, if we look at the uniquenesses:

```{r}
enframe(bem.15$uniquenesses, name="trait") %>% arrange(desc(value)) %>% left_join(traits)
```

some of these are still very high, so even 15 factors is not catching everything.

## time series