---
title: "March 6"
output: html_notebook
---

## Packages

Install `ggbiplot` below by installing `devtools` first, then:

```{r, eval=F}
library(devtools)
install_github("vqv/ggbiplot")
```


```{r}
library(ggbiplot) # biplots for discriminant analysis (and other things)
library(MASS) # for discriminant analysis (later)
library(tidyverse)
library(car)
library(lme4) # for mixed models
library(ggrepel) # for labelling points on plots
```

## Repeated measures (review)

slide 307



### Histamine in dogs

(repeated, faster)

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/dogs.txt"
dogs=read_table(my_url)
dogs
```

```{r}
response=with(dogs,cbind(lh0,lh1,lh3,lh5))
response
dogs.1=lm(response~drug,data=dogs)
```

We have some extra setup to express that these are the same thing measured at different times:

```{r}
times=colnames(response)
times.df=data.frame(times)
```

Then the analysis

```{r}
Manova(dogs.1, idata=times.df, idesign=~times)
```

The effect of drug is different at different times.

To investigate what, we can draw a graph. But for that, need *long* format:

```{r}
dogs %>% gather(timex,lh,lh0:lh5) %>% 
  mutate(time=parse_number(timex)) -> dogs.long
dogs.long
```

Make an interaction plot of log-histamine against time, labelled by drug:

```{r}
ggplot(dogs.long,aes(x=time, y=lh ,colour=drug, group=drug))+
  stat_summary(fun.y=mean,geom="point")+
  stat_summary(fun.y=mean,geom="line")
```

This shows that log-histamine levels are similar at time zero, but for all times after that, trimethaphan log-histamine levels are higher.

Spaghetti plot - join all the measurements for each individual (dog) with a line:

```{r}
ggplot(dogs.long,aes(x=time, y=lh, colour=drug, group=dog)) +
  geom_point()+geom_line()
```

Time zero really ought not to be included in the analysis, because there is no time for the drugs to have an effect. (There is usually a value noted at time zero so that we can confirm the drugs were not different then.) 

So let's go back to the original data frame and exclude time zero. This means redefining the response and following through:

```{r}
response=with(dogs,cbind(lh1, lh3, lh5)) # excluding time zero
dogs.2=lm(response~drug, data=dogs)
times=colnames(response)
times.df=data.frame(times)
Manova(dogs.2,idata=times.df, idesign=~times)
```

The interaction is no longer significant. Go back and look at interaction plot and spaghetti plot for why.

Cannot take it out (interaction of "within-subject" factor `times` and "between-subject" factor `drug`). So ignore it, because it is non-significant, and interpret the main effects: an effect of time, but an only marginally significant effect of drug. 


### Mixed model approach, using `lme4`:

This uses the "long format" `dogs.long`:

```{r}
dogs.long
```


Make sure to use the *text* time `timex` instead of the numerical time, to get it treated as categorical (otherwise gets treated as regression, linear in time, which it is not).

```{r}
dogs.3=lmer(lh~timex*drug+(1|dog), data=dogs.long)
drop1(dogs.3, test="Chisq")
```

This time the interaction is strongly significant.

or omit time zero:

```{r}
dogs.long %>% filter(time>0) -> d
dogs.4=lmer(lh~timex*drug+(1|dog), data=d) 
drop1(dogs.4, test="Chisq")
```

this time the interaction is still significant. (Maybe that drop from time 1 to time 3 is steeper for drug morphine.)

## Another example: exercise and diet

See slide 325.

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/exercise.txt"
exercise.long=read_tsv(my_url)
exercise.long
```

Diet and exercise type are between-subjects, but time is within-subjects. (Each subject does one diet and one type of exercise all the way through, but is measured at several different times.) 

Make *wide* for analysis:

```{r}
exercise.long %>% spread(time, pulse) -> exercise.wide
exercise.wide
```

Make response:

```{r}
response=with(exercise.wide, cbind(min01, min15, min30))
```

Predict from diet and exercise type:

```{r}
exercise.1=lm(response~diet*exertype, data=exercise.wide)
```

Manova:

```{r}
times=colnames(response)
times.df=data.frame(times)
Manova(exercise.1, idata=times.df, idesign=~times)
```

Now have significant *three*-way interaction: the effect of the combination of diet and exercise type is itself different at different times.

Spaghetti plot:

```{r}
ggplot(exercise.long, aes(x=time, y=pulse, group=id))+
  geom_point()+geom_line()
```

but normally we'd have a colour = something, and here we have two somethings: diet and exercise type. So instead have facets for these. With two things to facet by, use `facet_grid` to have one as rows and the other as columns:

```{r}
ggplot(exercise.long, aes(x=time, y=pulse, group=id))+
  geom_point()+geom_line()+
  facet_grid(diet~exertype)
```


(put the categorical variable with more levels on the $x$.)

Only for exercise type running is there any time effect worth noting. 

Can I do "simple effects" of diet for the runners? 

Pull out only the runners first:

```{r}
(exercise.wide %>%
  filter(exertype=="running") -> runners.wide)
```

and then redo all the same analysis on this data frame:

```{r}
response=with(runners.wide, cbind(min01, min15, min30))
runners.1=lm(response~diet, data=runners.wide)
times=colnames(response)
times.df=data.frame(times)
Manova(runners.1, idata=times.df, idesign=~times)
```

Even for the runners, the effect of time is still different for each diet.

Now that we are just focusing on one exercise type, we can do an interaction plot, for which we need long data:

```{r}
runners.wide
runners.wide %>% gather(time, pulse, min01:min30) -> runners.long
runners.long
```

the other way of doing this is to start with `exercise.long` and filter the runners:

```{r}
exercise.long %>% filter(exertype=="running")
```


and then the interaction plot:

```{r}
ggplot(runners.long, aes(x=time, y=pulse, colour=diet, group=diet))+
  stat_summary(fun.y=mean, geom="point") +
  stat_summary(fun.y=mean, geom="line")
```

These are not parallel: the difference in (mean) pulse rates gets larger as time passes. Is this real? Look at spaghetti plot:

```{r}
ggplot(runners.long, aes(x=time, y=pulse, colour=diet, group=id))+
  geom_point()+geom_line()
```

At 1 minute, the pulse rates are all mixed up, at 15 minutes the lowfat ones are mostly higher, and at 30 minutes they are all clearly higher. 


## Discriminant analysis

slide 340

notes on `select`1: slide 342

### Seed yields and weights

(again)

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/manova1.txt"
hilo=read_delim(my_url," ")
ggplot(hilo,aes(x=yield,y=weight,
  colour=fertilizer))+geom_point()
```

fertilizer had an effect on *combination* of yield and weight: if fertilizer was high, yield and weight were *both* high. This was significant: MANOVA:

```{r}
response=with(hilo,cbind(yield,weight))
hilo.2=lm(response~fertilizer,data=hilo)
Manova(hilo.2)
```

Discriminant analysis done with `lda` from `MASS`. Group is the "response", so flip around response and explanatory:

```{r}
hilo.1=lda(fertilizer~yield+weight,data=hilo)
hilo.1
```

`LD1` is a "score", made up by combining yield and weight, that best distinguishes high and low fertilizer. 

Here, both coefficients are negative, so:

- when yield and weight are both high, LD1 will be very *negative*.
- when yield and weight are both low, LD1 will be very *positive*.

LD1 maps the two variables into one dimension in a way that best distinguishes the groups.

How many LDs do you get? See slide 347. Here $\min(2, 2-1)=1$.

Each observation has a score on each LD, gotten by `predict`ing:

```{r}
p=predict(hilo.1)
p
```

(look at `class` and `posterior` shortly).

```{r}
cbind(hilo, p) %>% arrange(LD1) -> d
d
```

All the smallest (most negative) `LD1`s go with high fertilizer, and high yield and weight. All the most positive ones go with low fertilizer, and small yield and weight.

The LDs "reduce the dimension": here, LD1 says how the fertilizer groups differ in *one* dimension while yield and weight take two. So, can plot the first or first two LDs against the groups to see how the groups differ:

```{r}
ggplot(d, aes(x=fertilizer, y=LD1))+geom_boxplot()
```

High fertilizer goes with negative LD1, which goes with high yield and weight. Low fertilizer is the opposite.

More in `d`:

```{r}
d
```


`class` is the best guess at which fertilizer level each plant was at, based on its yield and weight. Tabulating with `fertilizer` shows how distinguishable the groups are:

```{r}
d %>% count(fertilizer, class)
```

All correct. The two fertilizer groups are very distinguishable based on yield and weight. 

Prettier:

```{r}
d %>% count(fertilizer, class) %>% 
  spread(class, n, fill=0)
```


The two `posterior` columns are how sure we are each plant was high or low fertilizer. Use `round` to avoid scientific notation:

```{r}
d %>% mutate_at(vars(starts_with("posterior")), ~round(.,3))
```

Most of these we are almost certain about. The 4th plant has some doubt: the yield is low, but the weight is high.



### Professions and leisure activities


```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/profile.txt"
active=read_delim(my_url," ")
active
active.1=lda(job~reading+dance+tv+ski,data=active)
```

```{r}
active.1
```

Most of the distinguishing is done by `LD1`. This depends (negatively) on attitudes towards dancing and TV-watching, `LD2` depends negatively on TV-watching.

```{r}
p=predict(active.1)
cbind(active, p) -> d
d
```

Plot individuals' scores by their actual job:

```{r}
ggplot(d, aes(x=x.LD1, y=x.LD2, colour=job, label=job))+geom_point()+geom_text_repel()
```

Or, having decided that things really only depend on attitudes towards dancing and TV-watching:

```{r}
ggplot(d, aes(x=dance, y=tv, colour=job, label=job))+geom_point()+geom_text_repel()
```

or, a biplot that shows both:

```{r}
ggbiplot(active.1, groups=factor(active$job))
```

The bellydancers like dancing, and the politicians don't really like dancing, but like watching TV.

Classification:

```{r}
d %>% count(job, class)
```

Everyone gotten right. Was this clear-cut?

```{r}
d %>% mutate_at(vars(starts_with("posterior")), ~round(., 3))
```

yep. Only the last one has any doubt at all.


Cross-validating the jobs data (all in one shot):

```{r}
active.2=lda(job~reading+dance+tv+ski, data=active, CV=T)
cbind(active, class=active.2$class, active.2$posterior) -> cv
cv
```

Correctness of predictions:

```{r}
with(cv, table(job, class))
```

One error this time. Why?

```{r}
ggplot(d, aes(x=x.LD1, y=x.LD2, colour=job, label=job))+geom_point()+geom_text_repel()
```

One of the bellydancers is a lot closer to the politicians. This shows up in the posterior probabilities:

```{r}
cv %>% mutate_at(vars(admin:politician), ~round(.,3))
```

The last bellydancer was "almost certainly" a politician! One of the administrators could also have been a politician.


